Enforced Invariant Confluence

### 2020-08-10 doc/the_vldb_journal_paper/rebuttal.tex

% Turn on the full reviewer comments if the turnonshowfeedback flag was passed
% in from the command line. This was taken from [1].
%
% [1]: https://stackoverflow.com/a/1466610/3187068
% \ifdefined\turnonshowfeedback
%   \toggletrue{showfeedback}
% \else
%   \togglefalse{showfeedback}
% \fi

\begin{document}
\begin{center}
  \ttlfnt Response to Reviewer Comments
\end{center}

\begin{multicols*}{2}
Below, you'll find our response to reviewer comments and our revisions. All
changes made to the paper are shown in \markrevisions{pink}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Overview}
Our original publication on \invariantconfluence{} is largely theoretical. We
provide new, and arguably improved, definitions of \invariantconfluence{}. We
characterize the relationship between \invariantclosure{} and
\invariantconfluence{}. We exploit the relationship to develop an
\invariantconfluence{} decision procedure. We invent the notion of
\mergereducibility{} and characterize its relationship with
\invariantclosure{} and \invariantconfluence{}. We invent the notion of
segmented \invariantconfluence{} and develop a segmented \invariantconfluence{}
decision procedure.

When our paper was invited to this journal, we took the opportunity to improve
the rigor and clarity of the core contributions of the paper: its theoretical
foundations. We added numerous proofs to verify the theorems in the paper, and
we added many examples and illustrations providing valuable intuition behind
the key results. We also added a significant amount of new material:
\begin{itemize}
  \item (Section 2.3)
    In our conference paper, we present a definition of invariant confluence
    that is slightly different than the original definition from Bailis et
    al.~\cite{bailis2014coordination}, but we do not comment on the difference.
    We added a subsection (Section 2.3) clarifying that the difference in
    definitions is superficial. We describe Bailis' original definition using
    our formalism and prove that the two definitions are equivalent.

@@

\paragraph{Reviewer 1: W2}
\begin{feedback}
  More experiments with some realistic workloads (like TPCC) to show the
  benefit of the proposed concepts in action are needed
\end{feedback}
We agree that having a number of realistic workloads in the paper would be
beneficial, but specifying, verifying, and explaining every workload requires a
significant amount of work for a marginal increase in the clarity of the paper.
The focus of the paper is the foundations, not the applications of these ideas.

\paragraph{Reviewer 1: W3}
\begin{feedback}
  Even though the automatic segmentation is left for future work, an example
  with a manual mapping of segments in a realistic workload (like TPCC) would
  show-case its potential
\end{feedback}
See above.

\paragraph{Reviewer 1: W4}
\begin{feedback}
  The main weak point of this specific submission is whether it has enough new
  material to to qualify for being an extended version
\end{feedback}
See our Overview section above.

\paragraph{Reviewer 1: D1}
\begin{feedback}
  W4 touches upon the ``new material'' for this extended version of the paper,
  and the above weak points W1, W2 and W3 can all serve as additional material.
\end{feedback}
See W1, W2, W3, and our Overview section above.

@@

  section 8.2 in example 12: ``and a optional'' $\to$ ``and an optional''
\end{feedback}
Fixed.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Reviewer 3}
\paragraph{Reviewer 3: Overview}
\begin{feedback}
  This manuscript is an extension of the VLDB 2019 conference version of
  ``Interactive Checks for Coordination Avoidance''. The original PVLDB paper
  is a solid piece of work that is well written and easy to follow. It was
  published in September 2018. In this manuscript, the authors elaborated more
  by adding more examples, proofs, and figures to make it easier for the reader
  to understand specific concepts and definitions. Some readers may find that
  the original paper is clear and enough without these additions. Moreover, the
  authors added two new sections, namely Section 4.3 and 7, which do not show a
  significant improvement to the original contribution. In almost two years, it
  was expected that the authors will address some of the future work mentioned
  in the original paper, such as ``In the future, we plan on exploring ways to
  help a user reason about unreachable states and ways to discover sets of
  unreachable states automatically.'' Overall, the authors did not manage to
  show that the additions were significant enough to warrant publication.
\end{feedback}
We address this feedback in detail below.

\paragraph{Reviewer 3: Section 2.1}
\begin{feedback}
  The authors revised a small part of this section to clarify that the merge
  operator of a distributed object does not have to satisfy any special
  mathematical properties like associativity and commutativity.
\end{feedback}
Yup. This was requested by many readers who assumed early on that the merge
operator did require some special properties. In Bailis et al.'s original
paper, for example, they assume the merge operator is associative, commutative,
and idempotent~\cite{bailis2014coordination}.

\paragraph{Reviewer 3: Section 2.3}
\begin{feedback}
  The authors added this section to prove that their definition of invariant
  confluence and the original definition from Bailis et al.\ are equivalent.
  The authors did not clarify why it is important to prove that both
  definitions are equivalent.
\end{feedback}
We believe that proving equivalent definitions are, in fact, equivalent is a
fundamental tenet of theoretical research. Without a proof of the two
definitions being equivalent, it's possible that our notion of
\invariantconfluence{} and Bailis et al.'s original notion differ.  If so, none
of our theorems or decision procedures could be applied to Bailis et al.'s
findings. We'd be talking about two completely different concepts.  Proving
they are equivalent unifies the work. Without this knowledge, proofs using the
different definitions would be mutually incompatible.

@@

\paragraph{Reviewer 3: Section 3}
\begin{feedback}
  The authors revised a small part of the original section to be in the form of
  a proof. For example in the PVLDB paper, the authors said: ``The forward
  direction of Theorem 2 follows immediately from Theorem 1. The backward
  direction holds because any two invariant satisfying states s1 and s2 must be
  reachable (by assumption), so their join s1 t s2 is also reachable. And
  because O is (s0, T, I)-confluent, all reachable points, including s1 t s2,
  satisfy the invariant.'' And in the VLDBJ submission, the authors said:
  ``Proof. The forward direction of Theorem 3 follows immediately from Theorem
  2. The backward direction holds because any two invariant satisfying states
  s1 and s2 must be reachable (by assumption), so their join s1 ⊔ s2 is also
  reachable. And because O is (s0, T, I)- confluent, all reachable points,
  including s1 ⊔s2, satisfy the invariant. Moreover s0 is reachable, so I(s0)
  because O is (s0,T,I)-confluent.''
\end{feedback}
Yup, a minor but (we hope) useful improvement.

\paragraph{Reviewer 3: Section 4.1}
\begin{feedback}
  the authors expanded their correctness proof of the interactive decision
  procedure and added a few other minor elaborations. The original proof in the
  PVLDB paper was clear enough to the reader.
\end{feedback}
We're happy that the original proof was clear to the reviewer, but we felt it
was not clear to most people.

@@

\paragraph{Reviewer 3: Section 4.2}
\begin{feedback}
  In Section 4.2, the authors added a very brief example showing why invariant
  closure is undecidable. The reader can find original section 4.2 is clear
  enough. Instead in both versions of section 4.2, i.e., PVLDB and VLDBJ, the
  authors said: ``In the future, we plan on exploring ways to help a user
  reason about unreachable states and ways to discover sets of unreachable
  states automatically.'' After almost two years of publishing the PVLDB
  version of this submission, this manuscript would be much stronger, if the
  authors explored these ways.
\end{feedback}
Automatically discovering unreachable states is hard, really hard. As we
describe in the paper, it is easy to verify that a state is reachable, simply
give an expression that evaluates to it. Verifying that a state is unreachable
is significantly harder, as we have to create a proof showing that there does
not exist any execution in which the state is reached. Open mathematical
problems like the Collatz conjecture~\cite{mottaanalysis} show that reasoning
about the reachability or unreachability of even simple programs can be
tremendously difficult. If we developed any substantial techniques to identify
unreachable states in a general way, we feel that would constitute a strong
publication on its own.

\paragraph{Reviewer 3: Section 4.3}
\begin{feedback}
  Section 4.3 is newly added to clarify that the decision procedure relies on
  human input to guide the decision procedure and explain how to mitigate these
  errors.
\end{feedback}
Yes, and we feel this is one of the core additions to the paper. The two
biggest drawbacks of interactivity are (1) the time and labor required of a
user and (2) the possibility of introducing errors. We break down the two types
of user errors (i.e.\ labeling an unreachable state as reachable and labeling
a reachable state as unreachable) and provide solutions for both, eliminating
the possibility of an erroneously verified \invariantconfluent{} object from
violating an invariant in deployment. This addition is very important for
anyone deploying \invariantconfluent{} objects in the wild. It's also
surprising and non-obvious. We originally did not think that our decision
procedures could be made tolerant to user errors.

@@

\paragraph{Reviewer 4: 1}
\begin{feedback}
  In the introduction the concept of ``invariant'' is missing some context. Where
  does this invariant come from? It looks like this invariant is a given
  invariant for the program executing on the replicated object, and it is not
  necessarily a tight invariant.  Clarifying this in advance would be useful,
  as I realized this only in the beginning of Section 2, I read this:
  ``Informally, a replicated object is invariant confluent *with respect to an
  invariant* if every replica of the object is guaranteed to satisfy the
  invariant despite the possibility of different replicas being concurrently
  modified or merged together.'' Mentioning the system model explicitly in the
  introduction also would help: ``a distributed object and accompanying
  invariant is replicated across a set of servers.''
\end{feedback}
We re-worded the introduction with your suggestion and also added a short
example of \invariantconfluence{} to the introduction.

\paragraph{Reviewer 4: 2}
\begin{feedback}
  Invariant confluence is about safety, however, could there be a liveness part
  of the problem worth investigating/discussing for replicated object
  management? Would it be possible that forgoing coordination across replicas
  could jeopardize/violate some liveness properties?
\end{feedback}
Forgoing coordination across replicas in favor of \invariantconfluence{}
actually \emph{increases} liveness, rather than jeopardizing or violating it.
Any alive replica of an \invariantconfluent{} object can service client
requests and execute transactions. This is the strongest form of liveness we
can hope to achieve. In fact, if we aim for strongly consistent replication,
then the CAP theorem tells us that total availability (the kind that
\invariantconfluence{} achieves) is impossible~\cite{brewer2012cap,
gilbert2002brewer}.

Note, however, that a \emph{segmented} \invariantconfluent{} object does suffer
from liveness problems, as we discuss in Section 6.4. Specifically,
transactions that require global coordination are live only as long as every
replica is live. We discuss this more in Reviewer 4: 5 below.

@@

\paragraph{Reviewer 4: 4}
\begin{feedback}
  Section 4 is about invariant tightening, which has a broader applicability in
  distributed systems/programs. Unfortunately the techniques used here are not
  new. Involving the user for identifying unreachable states in not practical,
  and the discussion on tolerating user errors is weak as it does not offer a
  remedy. Would using some heuristics and abstracting state provide any help
  here?
\end{feedback}
We agree that involving the user for identifying unreachable states is not
ideal, and in many cases not practical. We think our discussion of tolerating
user errors is also not ideal, but not weak. By requiring users to provide
proofs of reachability and by deploying the invariant $I - NR$ instead of $I$,
invariant violations are impossible.
%
Heuristics and abstracting state and other techniques from invariant tightening
would probably help, but they are out of scope of the paper. As you mention,
the techniques are not new. We view it as an orthogonal but complementary line
of research.

\paragraph{Reviewer 4: 5}
\begin{feedback}
  Section 6 on segmented invariant confluence is nice. It would be nice if
  Section 6.4 could be improved. I liked the fourth point in Section 6.4, would
  it be possible to elaborate on this. The sixth point however is bad, as
  replicating every server with state-machine-replication is overkill. Would it
  be possible to use use a membership service approach (as in Vertical Paxos
  and Chain Replication configuration) to prevent a single crash to flunk
  global coordination?
\end{feedback}
We added a couple paragraphs to Example 13 with a concrete example of the
optimization described in our fourth point. We also added a forward reference
in Section 6.4 to Example 13.

Your comment on our sixth point is interesting. We hadn't thought of that
before. Coincidentally, since our initial publication, we spent some time
researching membership changes in state machine replication protocols---which
you can find here:
\url{https://mwhittaker.github.io/publications/matchmaker_paxos.html}---so we
can answer your question with quite a bit of confidence. Assume a client sends
a transaction to a server $p$. The transaction happens to \emph{not} need
global coordination, so $p$ executes the transaction locally and sends back the
result to the client. This scenario makes membership changes difficult for the
following reason. Consider a different execution in which $p$ is partitioned
from the rest of the servers. The system erroneously thinks that $p$ has failed
and reconfigures to some other set of servers. However, $p$ and the client are
oblivious to this. The client can still contact $p$ and have transactions
executed, even though the system has moved to a completely new set of servers.
This can lead to situations that are impossible without membership changes. For
example, $p$ may be in a different active segment than the rest of the servers.

### 2020-08-10 doc/the_vldb_journal_paper/sections/evaluation.tex
Final tweaks.

\begin{revisions}
  For example, assume that $k = 3$ and that the three states are $s_1 = s_2 =
  s_3 = (p_1, p_2, p_3), (n_1, n_2, n_3) = (3, 4, 5), (3, 3, 3)$. Here, the
  value of the PN-counter is $(3 + 4 + 5) - (3 + 3 + 3) = 3$, which is
  non-negative as expected. If server 1 receives a decrement request, it
  \emph{cannot} execute the decrement without global coordination because doing
  so would cause $n_1$ to exceed the escrow threshold $k = 3$. Naively, it
  seems safe. Server 1 would transition to state $(3, 3, 5), (3, 3, 3)$ which
  has non-negative value $2$. However, if servers 2 and 3 concurrently perform
  decrements of their own, the three servers' states would merge to a negative
  value. Thus, the decrement requires global coordination to rule out the
  possibility of concurrent decrements.

  On the other hand, if server 2 receives a decrement request, it can safely
  execute it locally and transition to state $(3, 3, 5), (3, 3, 3)$. This is
  safe because server 2 knows that every server $i$ maintains $p_i \geq n_i$.
  Thus, after merging, the value of the PN-counter is $\sum_i p_i - \sum_i n_i
  = \sum_i (p_i - n_i)$ which is the sum of non-negative terms. Intuitively, if
  every server agrees to never go in debt, then the PN-counter in aggregate
  never goes in debt.
\end{revisions}

@@

  Escrow transactions are a good example where this optimization works
  particularly well. Assume that $k = 3$ and that the three states are $(3, 3,
  3), (3, 3, 3)$. Here, no server can process a decrement transaction locally.
  If a server receives a decrement transaction, it must execute it with global
  coordination. However, if the server instead defers the execution of the
  decrement transaction and happens to receive a number of increment
  transactions, then it can execute the decrement locally, without the need for
  global coordination.
\end{revisions}

>>>>

\begin{revisions}
  For example, assume that $k = 3$ and that the three states are $s_1 = s_2 =
  s_3 = (p_1, p_2, p_3), (n_1, n_2, n_3) = (3, 3, 3), (3, 2, 1)$. Here, the
  value of the PN-counter is $(3 + 3 + 3) - (3 + 2 + 1) = 3$, which is
  non-negative as expected. If server 1 receives a decrement request, it
  \emph{cannot} execute the decrement without global coordination because doing
  so would cause $n_1$ to exceed the escrow threshold $k = 3$. Naively, it
  seems safe. Server 1 would transition to state $(3, 3, 3), (4, 2, 1)$ which
  has non-negative value $2$. However, if servers 2 and 3 concurrently perform
  decrements of their own, the three servers' states would merge to a negative
  value. Thus, the decrement requires global coordination to rule out the
  possibility of concurrent decrements.

  On the other hand, if server 2 receives a decrement request, it can safely
  execute it locally and transition to state $(3, 3, 3), (3, 3, 1)$. This is
  safe because server 2 knows that every server $i$ maintains $p_i \geq n_i$.
  Thus, after merging, the value of the PN-counter is $\sum_i p_i - \sum_i n_i
  = \sum_i (p_i - n_i)$ which is the sum of non-negative terms. Intuitively, if
  every server agrees to never go in debt, then the PN-counter in aggregate
  never goes in debt.
\end{revisions}

@@

  Escrow transactions are a good example where this optimization works
  particularly well. Assume that instead of using PN-counters, we implement a
  replicated bank account balance as a tuple of timestamped integers $(x_1,
  x_2, \ldots, x_n)$, with the value of the bank account being the sum of the
  integers, with merging done element-wise, and with higher timestamps
  overriding lower timestamps~\cite{wu2019anna}. Every time a sever $s_i$
  increments or decrements its integer $x_i$, it increases the integer's
  timestamp. We consider the segmentation in which decrements do not require
  global coordination so long as every individual integer remains non-negative.

  If a server's integer is zero, it cannot process a decrement transaction
  locally, it must execute it with global coordination. However, if the server
  instead defers the execution of the decrement transaction and happens to
  receive a number of increment transactions, then it can execute the decrement
  locally, without the need for global coordination.
\end{revisions}
